### 2014-11-04 thesis/method/classification.tex
Final version of thesis

Concretely, given a feature
vector of the form \textit{(age, gender, has\_cancer)},
1R will split the ages of
all patients into discrete categories
(such as \textit{0-20, 21-40, 41-65, 66+}). It
will then check the class labels for all training examples in the data set
that have \textit{age} in the 0-20 category
(as that is the first value for the discretised \textit{age} feature) and
note the most common class, proceeding similarly for the rest of the
categories of \textit{age}. Then it will consider \textit{gender}, which has
two possible values $\{$\textit{male, female}$\}$, and finally \textit{has\_cancer}
with two values $\{$\textit{true, false}$\}$. This results in 8 rules,
corresponding to three features.
An example of a rule is that \textit{age 0-20} most commonly
has ``LOS $\leq$ 2 days''. For each of these rules, 1R will attempt to
use only that rule to predict the LOS class for each example in the data set,
noting down the error rate. 1R then chooses the feature that produces rules
with the highest total accuracy.

\section{Na\"{i}ve Bayes}
Unlike 1R, which considers only one feature in the feature vector, Na\"{i}ve
Bayes (NB) classifiers allows all features to contribute equally to the
classification decision. Such classifiers apply Bayes' theorem with the
(unrealistic) assumption of independence between features, hence the
description \textit{na\"{i}ve}. This means that we
calculate the probability that a particular feature value belongs in a class
without considering the effect of other features on this probability. This
probability can be computed by
counting the frequency with which the value is associated with a class in the
training data. After these probabilities are computed, classifying a feature
vector simply involves multiplying the probabilities of each feature belonging
to each possible class. The predicted class for that feature vector is then the
class with the highest probability.

Speaking mathematically, Bayes' theorem states that:
\begin{equation*}
\mathbb{P}(Class_i|F_1,F_2,\ldots,F_n) =
  \dfrac{\mathbb{P}(F_1,F_2,\ldots,F_n|Class_i)\mathbb{P}(Class_i)}{\mathbb{P}(F_1,F_2,\ldots,F_n)}
\end{equation*}
Since we assume that the features are independent, we can simplify the above as:
\begin{equation*}
\mathbb{P}(Class_i|F_1,F_2,\ldots,F_n) =
\dfrac{\mathbb{P}(F_1|Class_i)\mathbb{P}(F_2|Class_i)\ldots \mathbb{P}(F_n|Class_i)\mathbb{P}(Class_i)}{\mathbb{P}(F_1,F_2,\ldots,F_n)}
\end{equation*}
This gives us the probability that the feature vector $(F_1,F_2,\ldots,F_n)$
belongs in the $i$-th class. Classifying the feature vector involves
calculating this probability for each for each class and choosing the class
with the highest probability.

Since we compute the probability of each feature value belonging in a
particular class by counting the frequency of each value's occurrence with
respect to the class, it is sometimes possible that a feature value has zero
probability of being in a certain class. This would make the numerator of the
above equation zero. A feature vector that could otherwise have been highly
likely to belong in class $i$ is now counted as having zero probability of
belonging in the class. In order to ensure that each probability is non-zero,
a small constant is added to all probabilities in a process called
\textit{Laplace smoothing}, which is commonly used for NB classifiers.

@@

The only thing left to specify
now is how to decide which attribute to split on. Note that when splitting,
we would like to create subsets that are as ``pure''
-- that is, containing as many
instances of the same class -- as possible, so that the splitting may
terminate earlier. Since we would like our trees to be as small as possible to
prevent overfitting and allow the best generalisation ability,
we would therefore like to choose the feature to split on such that the
created subsets were as pure as possible. The measure of purity used in
learning decision trees is the \textit{information}, and is measured in
\textit{bits}. At any node (or subset) of the tree, this measure represents
the expected amount of information that would be needed to classify a new
example, given that the example reached the node. The information at each
node is calculated using the \textit{entropy} function:
\begin{equation}
\label{eqn:entropy}
\mathrm{entropy}(p_1,p_2,\ldots,p_n) = -p_1\mathrm{log}p_1 -p_2\mathrm{log}p_2 \ldots -p_n\mathrm{log}p_n
\end{equation}
Given $k$ classes, $c_1,\ldots,c_k$ is the number of instances in each
class at a node, and $\sum_{i=1}^k c_i = c$ is the total number of instances
at that node, the information of that node is then:
\begin{equation*}
\mathrm{info}(c_1,\ldots,c_k) = \mathrm{entropy}\left(\frac{c_1}{c},\frac{c_2}{c},\ldots,\frac{c_k}{c}\right)
\end{equation*}

@@

Decision trees are fast to train and produce models that are straightforward
for humans to interpret, the latter of which is particularly valuable in a
clinical environment. We will use the C4.5 tree induction algorithm, which is
considered the state-of-the-art and handles different types of feature values
as well as missing values.

\section{Logistic Regression}
Logistic regression (LR) is a regression technique used for when the dependent
variable is dichotomous or two-valued, and thus it is a suitable candidate to
use for our LOS classification problem.
Previous work on trauma LOS prediction by
Dinh et. al. \cite{Dinh2013a}, as well as many attempts at modelling the LOS
in various medical domains, have all used logistic regression, often as a
starting point.

For a class variable with two possible values that occur with
probability $p$ and $1-p$ respectively, and a feature vector
$(F_1,F_2,\ldots,F_n)$, the logistic regression model is:
\begin{equation*}
\begin{aligned}
\mathrm{logit}(p) &= \mathrm{log} \left(\dfrac{p}{1-p}\right) \\
  &= \beta_0 + \beta_1F_1 + \beta_2F_2 + \ldots + \beta_nF_n
\end{aligned}
\end{equation*}
For LOS classification, this can be rearranged to:
\begin{equation*}
\begin{aligned}
p &= \mathbb{P}(Class=\mathrm{LOS }\leq2\mathrm{ days}|F_1,F_2,\ldots,F_n) \\
  &= \dfrac{e^{\beta_0 + \sum^n_{i=1}\beta_i F_i}}{1 + e^{\beta_0 + \sum^n_{i=1}\beta_i F_i}}
\end{aligned}
\end{equation*}

Training the logistic regression classifier, then, is a matter of finding the
$\beta_i$ that will maximise the likelihood (conditional probability of the
data given the values for $\beta_i$) of the given data set. This is done using
numerical methods, where estimates of the $\beta_i$ are chosen, the likelihood
computed, and the $\beta_i$ are updated, and so on: this process is repeated
until parameter estimates converge or do not change more than a given threshold
value.

@@

Given a set of $n$-feature vectors, we can represent each vector as a point in
$n$-dimensional space, where each feature is a dimension. The SVM will try to
find a \textit{hyperplane} that separates the two classes. Usually there are
many such hyperplanes; the distinguishing aspect of an SVM is that it will
find the hyperplane that maximises the distance between the two nearest points
in each class. This distance is called the \textit{margin}, and the intuition
is that maximising it will give us the greatest possible separation between the
classes -- the classifier comes no closer to either class than it needs to.
This gives the SVM the best generalising ability to classify new examples.
\begin{figure}[h]
\includegraphics{images/method/svm-example.pdf}
\caption{An example of a 2-dimensional SVM. The solid line indicates the
maximum margin hyperplane.}
\label{fig:svm}
\end{figure}

Figure \ref{fig:svm} shows an example of a two-dimensional feature space, with
two classes of points (black and white). We can see the hyperplane, in this
case a line since we are only in two dimensions, separating the two classes
with as much margin as possible. In
this case, the points are linearly separable: all white points lie on one side
of the hyperplane, and all black points lie on the other. In practice, this is
not common, and often data cannot be separated cleanly by a hyperplane.
Note that the points closest to the margin are those that are most likely to be
ambiguous when the data cannot be cleanly separated. The SVM will progressively
remove more and more of these ambiguous points from consideration when it
cannot find a hyperplane until it finds a suitable one.

Instead of removing points from consideration by the SVM, we can also specify
a \textit{kernel function}, which transforms the original feature space into
some other non-linear space. Commonly, polynomial functions of degree 2 or
greater are used, and theoretically they can approximate non-linear boundaries
between classes to any arbitrary accuracy. However, this is usually not
practically possible with data sets containing even a modest number of features
(around 10), due to the computational complexity of estimating a large number
of coefficients that are introduced by the transformation.

SVMs can be specified mathematically as a type of optimisation problem known
as \textit{constrained quadratic optimisation}. Given $m$ training vectors
$\mathbf{x_i} \in \mathbb{R}^n, i=1,\ldots,m$ and their labels
$y_i \in \{-1,+1\}, i=1,\ldots,m$, the optimisation problem can be written as:
\begin{equation*}
\begin{aligned}
& \underset{\mathbf{w},b}{\mathrm{minimise}}
  && ||\mathbf{w}|| + C\sum_{i=1,\ldots,m} s_i \\
& \text{subject to}
  && y_i(\mathbf{w}^T\phi(\mathbf{x_i})+b) \geq 1 \\
& && s_i \geq 0, i = 1,\ldots,m
\end{aligned}
\end{equation*}
where $\phi(\mathbf{x})$ is the kernel function, and
$\mathbf{w}\cdot\mathbf{x_i} + b = 0$ is the hyperplane that the SVM finds.
The $s_i$'s allow the SVM to ignore some examples if it cannot find a suitable
hyperplane, and $C$ (sometimes called the SVM's complexity constant) is a
constant specified by the user that controls how much we want to balance
finding the best possible margin with allowing misclassified examples.
Note that for a linear kernel, $\phi(\mathbf{x}) = \mathbf{x}$. Training the
SVM involves solving this optimisation problem and obtaining values for
$\mathbf{w}$ and $b$.

@@

\begin{figure}[h]
\centering
\includegraphics{images/method/mlp.pdf}
\caption{An MLP with one hidden layer. There are $n$ input nodes, $k$ nodes
in the hidden layer, and one output node.}
\label{fig:mlp}
\end{figure}

MLPs can be trained in a number of ways, the most common of which is
\textit{backpropagation}. Suppose that we have the network above, with one
input, hidden and output layer respectively. From the perceptron learning
algorithm, we can update the weights of the hidden neurons leading to the
output neuron, but we do not know what the outputs of these hidden neurons are.
Backpropagation solves this problem by updating the weights of the various
connections to the hidden units based on the strength of the unit's
contribution to the final prediction, using a mathematical optimisation
algorithm called \textit{gradient descent}. This requires taking derivatives,
so the threshold function we used for the perceptron is usually replaced with
the sigmoid function $f(x) = \frac{1}{1+e^{-x}}$. We also define an error
function $E=\frac{1}{2}(\text{actual} - \text{predicted})^2$ -- the one most
commonly used -- which is
minimised in the gradient descent process. Updating the weights for each
connection in the network now involves finding the value of the derivative,
multiplying it by a small constant between 0 and 1 called the
\textit{learning rate}, and subtracting this value from the current weight
value, until the minimum for the error function is reached. Note that the
error function is quadratic, so it is guaranteed to have a minimum. The
updating of the weights is continued until the minimum is reached or if we
are ``close enough'' to the minimum.

@@

Like the Na\"{i}ve Bayes classifier, $k$-NN allows all features to be taken
into account equally when classifying a new example. However, since feature
values often span different ranges and units of measurement, a particularly
large range of values for a feature would contribute much more to the distance
than another feature that spans a lower set of values.
To avoid this, it is important to \textit{normalise} the
values of each numeric feature before using $k$-NN: for each feature $i$, we
calculate the normalised value of the feature for each training example. This
is done using the following relationship:
\begin{equation*}
a_i = s\left(\dfrac{v_i - \mathrm{min }v_i}{\mathrm{max }v_i - \mathrm{min }v_i}\right) + t
\end{equation*}
where $a_i$ is the normalised value, $v_i$ is the actual value in the data
set, max $v_i$ and min $v_i$ are taken over all training examples, $s$ is a
scaling factor and $t$ indicates how much to translate the range by. If $s=1$
and $t=0$, then the range of normalised values will be in $[0,1]$.

\subsection{Existing and Proposed Extensions}
There have been many extensions proposed to the $k$-NN algorithm described
above. These usually seek to find a more suitable distance metric for the
problem domain at hand, or to reduce the storage requirements of the original
$k$-NN algorithm. We will focus on two extensions to the Euclidean distance:
K* and our proposed algorithm, Ranked Distance.

@@

We use K* as a comparison to the basic NN algorithm and also as a comparison
to our proposed Ranked Distance Nearest Neighbours algorithm.

\subsubsection{Proposed Approach: Ranked Distance Nearest Neighour}
\label{sec:rankeddist}
\paragraph{Intuition}
As we mentioned earlier, all features are considered equally in a $k$-NN
classifier, which implies that all features are equally important in predicting
the class. In practice, however, this is not always the case: to predict the
LOS of a trauma patient, the severity of their injury should affect the final
LOS more than whether or not they can speak English (which could be just two
of many features that are recorded about a patient). Our contribution is a
new NN algorithm that uses a modified distance function which takes into
account the relative importance of all features called Ranked Distance
Nearest Neighbour (RD).

@@

\paragraph{Assignment of weights}
The weights $w_i$ can be tuned to match the particular problem that is being
investigated. For our LOS classification problem, we will consider one way
of weighting features.
First, we compute the correlation coefficients using:
\begin{equation*}
r_i = \dfrac{\sum_{i=1}^m (x_i-\bar{x})(c-\bar{c})}{\sqrt{(\sum_{i=1}^m x_i-\bar{x})(\sum_{i=1}^m c-\bar{c})}}
\end{equation*}
where $m$ is the number of training examples, $x_i$ is the value of feature
$i$, and $\bar{x}$ and $\bar{c}$ are the arithmetic averages of the feature
values and the class values respectively. 
This gives us a ranking of the importance of each feature to the prediction of
the class, with 1 being the highest rank (indicating the most importance) and
$n$ being the lowest rank (indicating the least importance).
Instead of using the values of the correlation directly to weight
the distance contribution of each feature, we specify a function:
\begin{equation*}
f : \mathrm{Rank} \rightarrow \mathbb{R}, \mathrm{Rank} \in \{1,2,\ldots,n\}
\end{equation*}
that describes
how the weights of the features vary with the rank. This function should
inituitively be non-increasing and should produce lower values when the rank
number is greater, meaning that less important features have a lower weight
in the distance calculation.
Note that choosing $f(\mathrm{Rank}) = k$ for any non-zero constant $k$
results in all features contributing equally to the distance calculation.
We will use $f(\mathrm{Rank}) = \frac{1}{\mathrm{Rank}^c}$, where
$c$ varies from 0 to 1.

\section{Summary}
In this chapter we have outlined all of the learning algorithms that we will
be evaluating with respect to our classification problem, from the simple
Zero-R and 1R to Na\"{i}ve Bayes, decision trees, logistic regression,
$k$-nearest neighbour (and K*), to the more
sophisticated support vector machines and multi-layer perceptrons. We also
described our proposed new algorithm, Ranked Distance Nearest Neighbours.
The reason
for our inclusion of so many algorithms is two-fold. Firstly, many of
these algorithms have not been extensively applied in various medical settings
to the problem of LOS prediction. Using a wide range of algorithms from various
paradigms will enable us to see what algorithms perform better than others.
Secondly, evaluating several classifiers will allow us to make comparisons
between different feature selection methods that are independent of the
classifier and give us more insight into the nature of the data set and of the
problem.

>>>>

Concretely, given a feature
vector of the form \textit{(age, gender, has\_cancer)},
1R will split the ages of
all patients into discrete categories
(such as \textit{0-20, 21-40, 41-65, 66+}). It
will then check the class labels for all training examples in the data set
that have \textit{age} in the 0-20 category
(as that is the first value for the discretised \textit{age} feature) and
note the most common class, proceeding similarly for the rest of the
categories of \textit{age}. Then it will consider \textit{gender}, which has
two possible values $\{$\textit{male, female}$\}$, and finally \textit{has\_cancer}
with two values $\{$\textit{true, false}$\}$. This results in 8 rules,
corresponding to three features.
An example of a rule is that \textit{age 0-20} most commonly
has ``LOS $\leq$ 2 days''. For each of these rules, 1R will attempt to
use only that rule to predict the LOS class for each example in the data set,
noting down the error rate. 1R then chooses the feature that produces rules
with the highest total accuracy across all discretised feature values.

\section{Na\"{i}ve Bayes}
Unlike 1R, which considers only one feature in the feature vector, Na\"{i}ve
Bayes (NB) classifiers allows all features to contribute equally to the
classification decision. Such classifiers apply Bayes' theorem with the
(unrealistic) assumption of independence between features, hence the
description \textit{na\"{i}ve}. This means that we
calculate the probability that a particular feature value belongs in a class
without considering the effect of other features on this probability. This
probability can be computed by
counting the frequency with which a feature value is associated with a class in the
training data. After these probabilities are computed, classifying a feature
vector simply involves multiplying the probabilities of each feature belonging
to each possible class. The predicted class for that feature vector is then the
class with the highest probability.

Speaking mathematically, Bayes' theorem states that:
\begin{equation*}
\mathbb{P}(Class_i|F_1,F_2,\ldots,F_n) =
  \dfrac{\mathbb{P}(F_1,F_2,\ldots,F_n|Class_i)\mathbb{P}(Class_i)}{\mathbb{P}(F_1,F_2,\ldots,F_n)}
\end{equation*}
Since we assume that the features are independent, we can simplify the above as:
\begin{equation*}
\mathbb{P}(Class_i|F_1,F_2,\ldots,F_n) =
\dfrac{\mathbb{P}(F_1|Class_i)\mathbb{P}(F_2|Class_i)\ldots \mathbb{P}(F_n|Class_i)\mathbb{P}(Class_i)}{\mathbb{P}(F_1,F_2,\ldots,F_n)}
\end{equation*}
This gives us the probability that the feature vector $(F_1,F_2,\ldots,F_n)$
belongs in the $i$-th class. Classifying the feature vector involves
calculating this probability for each class and choosing the class
with the highest probability.

Since we compute the probability of each feature value belonging in a
particular class by counting the frequency of each value's occurrence with
respect to the class, it is sometimes possible that a feature value has zero
probability of being in a certain class. This would make the numerator of the
above equation zero. A feature vector that could otherwise have been highly
likely to belong in class $i$ is then counted as having zero probability of
belonging in the class. In order to ensure that each probability is non-zero,
a small constant is added to all probabilities in a process called
\textit{Laplace smoothing}, which is now commonly used for NB classifiers.

@@

We now only need to specify how to decide which attribute to split on at
each step. Note that when splitting,
we would like to create subsets that are as ``pure''
-- that is, containing as many
instances of the same class -- as possible, so that the splitting may
terminate earlier. Since we would like our trees to be as small as possible to
prevent overfitting and allow the best generalisation ability,
we would therefore like to choose the feature to split on such that the
created subsets were as pure as possible. The measure of purity used in
learning decision trees is called \textit{information}, and is measured in
\textit{bits}. At any node (or subset) of the tree, this measure represents
the expected amount of information that would be needed to classify a new
example, given that the example reached the node. The information at each
node is calculated using the \textit{entropy} function:
\begin{equation}
\label{eqn:entropy}
\mathrm{entropy}(p_1,p_2,\ldots,p_n) = -p_1\mathrm{log}p_1 -p_2\mathrm{log}p_2 \ldots -p_n\mathrm{log}p_n
\end{equation}
Given $k$ classes, $c_1,\ldots,c_k$ is the number of instances in each
class at a node, and $\sum_{i=1}^k c_i = c$ is the total number of instances
at that node, the information of that node is then:
\begin{equation*}
\mathrm{info}(c_1,\ldots,c_k) = \mathrm{entropy}\left(\frac{c_1}{c},\frac{c_2}{c},\ldots,\frac{c_k}{c}\right)
\end{equation*}

@@

Decision trees are fast to train and produce models that are straightforward
for humans to interpret, the latter of which is particularly valuable in a
clinical environment. We will use the C4.5 tree induction algorithm
\cite{Quinlan1993}, which is considered the state-of-the-art in decision trees
and handles different types of feature values as well as missing values.

\section{Logistic Regression}
Logistic regression (LR) is a regression technique used for when the dependent
variable is dichotomous or two-valued, and thus it is a suitable candidate to
use for our LOS classification problem.
Previous work on trauma LOS prediction by
Dinh et. al. \cite{Dinh2013a} as well as many attempts at modelling the LOS
in various medical domains have all used logistic regression, often as a
starting point.

For a class variable with two possible values that occur with
probability $p$ and $1-p$ respectively, and a feature vector
$(F_1,F_2,\ldots,F_n)$, the logistic regression model is:
\begin{equation*}
\begin{aligned}
\mathrm{logit}(p) &= \mathrm{log} \left(\dfrac{p}{1-p}\right) \\
  &= \beta_0 + \beta_1F_1 + \beta_2F_2 + \ldots + \beta_nF_n
\end{aligned}
\end{equation*}
For our LOS classification problem, this can be rearranged to:
\begin{equation*}
\begin{aligned}
p &= \mathbb{P}(Class=\mathrm{LOS }\leq2\mathrm{ days}|F_1,F_2,\ldots,F_n) \\
  &= \dfrac{e^{\beta_0 + \sum^n_{i=1}\beta_i F_i}}{1 + e^{\beta_0 + \sum^n_{i=1}\beta_i F_i}}
\end{aligned}
\end{equation*}

Training the logistic regression classifier, then, is a matter of finding the
$\beta_i$ that will maximise the likelihood (conditional probability of the
data given the values for $\beta_i$) of the given data set. This is done using
numerical methods, where estimates of the $\beta_i$ are chosen, the likelihood
computed, the $\beta_i$ updated accordingly, and so on: this process is repeated
until parameter estimates converge or do not change more than a given threshold
value.

@@

Given a set of $n$-feature vectors, we can represent each vector as a point in
$n$-dimensional space, where each feature is a dimension. The SVM will try to
find a \textit{hyperplane} that separates the two classes. Usually there are
many such hyperplanes; the distinguishing aspect of an SVM is that it will
find the hyperplane that maximises the distance between the two nearest points
in each class. This distance is called the \textit{margin}, and the intuition
is that maximising it will give us the greatest possible separation between the
classes -- the classifier comes no closer to either class than it needs to.
This gives the SVM the best generalising ability to classify new examples.
\begin{figure}[h]
\includegraphics{images/method/svm-example.pdf}
\caption{An example of a 2-dimensional SVM. The solid line indicates the
maximum margin hyperplane, and the dotted line indicates the margin.}
\label{fig:svm}
\end{figure}

Figure \ref{fig:svm} shows an example of a two-dimensional feature space, with
two classes of points (black and white). We can see the hyperplane, in this
case a line since we are only in two dimensions, separating the two classes
with as much margin as possible. In
this case, the points are linearly separable: all white points lie on one side
of the hyperplane, and all black points lie on the other. In practice, this is
not common, and often the data cannot be separated cleanly by a hyperplane.
Note that the points closest to the margin are those that are most likely to be
ambiguous when the data cannot be cleanly separated. If it cannot find a
hyperplane, the SVM will progressively
remove more and more of these ambiguous points from consideration 
until it finds a suitable one.

Instead of removing points from consideration by the SVM, we can also specify
a \textit{kernel function}, which transforms the original feature space into
some other non-linear space. Commonly, polynomial functions of degree 2 or
greater are used, and theoretically they can approximate non-linear boundaries
between classes to any arbitrary accuracy. However, this is usually not
practically possible with data sets containing even a modest number of features
(around 10) due to the computational complexity of estimating a large number
of coefficients that are introduced by the transformation.

SVMs can be specified mathematically as a type of optimisation problem known
as \textit{constrained quadratic optimisation}. Given $m$ training vectors
$\mathbf{x_i} \in \mathbb{R}^n, i=1,\ldots,m$ and their labels
$y_i \in \{-1,+1\}, i=1,\ldots,m$, the optimisation problem can be written as:
\begin{equation*}
\begin{aligned}
& \underset{\mathbf{w},b}{\mathrm{minimise}}
  && ||\mathbf{w}|| + C\sum_{i=1,\ldots,m} s_i \\
& \text{subject to}
  && y_i(\mathbf{w}^T\phi(\mathbf{x_i})+b) \geq 1 \\
& && s_i \geq 0, i = 1,\ldots,m
\end{aligned}
\end{equation*}
where $\phi(\mathbf{x})$ is the kernel function, and
$\mathbf{w}\cdot\mathbf{x_i} + b = 0$ is the hyperplane that the SVM finds.
The $s_i$ allow the SVM to ignore some examples if it cannot find a suitable
hyperplane, and $C$ (sometimes called the SVM's complexity constant) is a
constant specified by the user that controls how much we want to balance
finding the best possible margin with allowing misclassified examples.
Note that for a linear kernel, $\phi(\mathbf{x}) = \mathbf{x}$. Training the
SVM involves solving this optimisation problem and obtaining values for
$\mathbf{w}$ and $b$, which define the hyperplane.

@@

\begin{figure}[h]
\centering
\includegraphics{images/method/mlp.pdf}
\caption{An MLP with one hidden layer. There are $n$ input nodes, $k$ nodes
in the hidden layer, and one output node which contains the prediction of
the network.}
\label{fig:mlp}
\end{figure}

MLPs can be trained in a number of ways, the most common of which is
\textit{backpropagation}. Suppose that we have the network above, with one
input, hidden and output layer respectively. From the perceptron learning
algorithm, we can update the weights of the hidden neurons leading to the
output neuron, but we do not know what the outputs of these hidden neurons are.
Backpropagation solves this problem by updating the weights of the various
connections to the hidden units based on the strength of the unit's
contribution to the final prediction, using a mathematical optimisation
algorithm called \textit{gradient descent}. This requires taking derivatives,
so the threshold function we used for the perceptron is usually replaced with
the sigmoid function $f(x) = \frac{1}{1+e^{-x}}$. We also define an error
function $E=\frac{1}{2}(\text{actual} - \text{predicted})^2$ -- the one most
commonly used in practice -- which is
minimised in the gradient descent process. Updating the weights for each
connection in the network now involves finding the value of the derivative,
multiplying it by a small constant between 0 and 1 called the
\textit{learning rate}, and subtracting this value from the current weight
value, until the minimum for the error function is reached. Note that the
error function is quadratic, so it is guaranteed to have a minimum. The
updating of the weights is continued until the minimum is reached or if we
are ``close enough'' to the minimum.

@@

Like the Na\"{i}ve Bayes classifier, $k$-NN allows all features to be taken
into account equally when classifying a new example. However, since feature
values often span different ranges and units of measurement, a particularly
large range of values for a feature would contribute much more to the distance
than another feature that spans a lower set of values.
To avoid this, it is important to \textit{normalise} the
values of each numeric feature before using $k$-NN: for each feature $i$, we
calculate the normalised value of the feature for each training example. This
is done using the following relationship:
\begin{equation*}
a_i = s\left(\dfrac{v_i - \mathrm{min }v_i}{\mathrm{max }v_i - \mathrm{min }v_i}\right) + t
\end{equation*}
where $a_i$ is the normalised feature value, $v_i$ is the actual value in the
data set, max $v_i$ and min $v_i$ are taken over all training examples, $s$ is
a scaling factor and $t$ indicates how much to translate the range by. If $s=1$
and $t=0$, then the range of normalised values will be in $[0,1]$.

\subsection{Existing and Proposed Extensions}
There have been many extensions proposed to the $k$-NN algorithm described
above. These usually seek to find a more suitable distance metric for the
problem domain at hand, or to reduce the storage requirements of the original
$k$-NN algorithm (since the classifier makes no attempt to generalise from the
examples, all of them must be stored).
We will focus on two algorithms that extend the standard Euclidean distance:
K* and our proposed approach, Ranked Distance Nearest Neighbour.

@@

We use K* as a comparison to the basic NN algorithm and also as a comparison
to our proposed Ranked Distance Nearest Neighbour algorithm. Additionally,
K* has not been yet been applied in predicting LOS, and its performance in this
domain is worth investigating.

\subsubsection{Proposed Approach: Ranked Distance Nearest Neighour}
\label{sec:rankeddist}
\paragraph{Intuition}
As we mentioned earlier, all features are considered equally in a $k$-NN
classifier, which implies that all features are equally important in predicting
the class. In practice, however, this is not always the case: for example,
to predict the
LOS of a trauma patient, the severity of their injury should affect the final
LOS more than whether or not they can speak English (which could be just two
of many features that are recorded about a patient). Our contribution is a
new NN algorithm, Ranked Distance Nearest Neighbour (RD), which uses a modified
distance function that takes into account the relative importance of each
feature in predicting the class.

@@

\paragraph{Assignment of weights}
The weights $w_i$ can be tuned to match the particular problem that is being
investigated. For our LOS classification problem, we will consider one way
of weighting features.
First, we compute the correlation coefficients using:
\begin{equation*}
r_i = \dfrac{\sum_{i=1}^m (x_i-\bar{x})(c-\bar{c})}{\sqrt{(\sum_{i=1}^m x_i-\bar{x})(\sum_{i=1}^m c-\bar{c})}}
\end{equation*}
where $m$ is the number of training examples, $x_i$ is the value of feature
$i$, and $\bar{x}$ and $\bar{c}$ are the arithmetic averages of the feature
values and the class values respectively. 
This gives us a ranking of the importance of each feature to the prediction of
the class, with 1 being the highest rank (indicating the most importance) and
$n$ being the lowest rank (indicating the least importance).
Instead of using the values of the correlation directly to weight
the distance contribution of each feature, we specify a function:
\begin{equation*}
f : \mathrm{Rank} \rightarrow \mathbb{R}, \text{ }\mathrm{Rank} \in \{1,2,\ldots,n\}
\end{equation*}
that describes
how the weights of the features vary with the rank. This function should
inituitively be non-increasing: that is, it should produce lower values when
the rank number is greater (which indicates a lower rank), meaning that less
important features have a lower weight in the final distance calculation.
Note that choosing $f(\mathrm{Rank}) = k$ for any non-zero constant $k$
results in all features contributing equally to the distance calculation.
We will use $f(\mathrm{Rank}) = \dfrac{1}{\mathrm{Rank}^c}$, where
$c$ varies from 0 to 1.

\section{Summary}
In this chapter we have outlined all of the learning algorithms that we will
be evaluating with respect to our classification problem, from the simple
Zero-R and 1R to Na\"{i}ve Bayes, decision trees, logistic regression,
$k$-nearest neighbour (and K*), to the more
sophisticated support vector machines and multi-layer perceptrons. We also
described our proposed new algorithm, Ranked Distance Nearest Neighbour.
The reason
for our inclusion of so many algorithms is two-fold. Firstly, many of
these algorithms have not been extensively applied in various medical settings
to the problem of LOS prediction. Using a wide range of algorithms from various
paradigms will enable us to see what algorithms perform better than others.
Secondly, evaluating several classifiers will allow us to make comparisons
between different feature selection methods that are independent of the
classifier and give us more insight into the nature of the data set and of the
problem itself.
